### 2007-01-15 Documentation/part2.tex
JK - Minor corrections in chapter 11

In this chapter, we give a detailed definition of the \AEDL\ features. We also show the underlying mathematical concepts, where applicable. We do not exactly follow the separation of source, oscillation, and detection properties, since most issues more or less involve the detection. Instead,
we illustrate many of the features of the \GLOBES\ simulation successively
in the logical order of their definition, and demonstrate how they translate into \AEDL .

@@

As we have disussed before, \GLOBES\ can only deal with point sources. Thus,  it is not possible to study effects from the finite size of the neutrino production region, such as in the sun or in reactor experiments with many
neutrino sources (\eg, KamLAND). Therefore, a neutrino source in \GLOBES\ can, in general, be characterized by the flux spectrum for each neutrino flavor, the CP sign (neutrinos or antineutrinos), and the total luminosity
of the source.

@@

\index{aedl}{nuflux@{\tt nuflux}!time@{\tt "@time}}
There are two principal ways to initialize a neutrino flux: 
Either one can use 
a built-in source, or one can provide a file. In both cases,
a flux is defined by the environment {\tt nuflux}, such as
\index{aedl}{nuflux@{\tt nuflux}}
\begin{quote}
  {\tt nuflux(\#name)<\\
\tb $\ldots$\\
\tb @time = 8.0 \\
>}
\end{quote}
with a running time of $8$ years. Note that the running time is used within
the {\tt nuflux} environment. This feature can be used to load the neutrino and antineutrino fluxes separately, in order to combine them with different
running times within one experiment. The name of the flux {\tt \#name} will later be referred to in the channel definitions. Note that \GLOBES\ versions older than 3.0 use the still supported {\tt flux} environment, which is different from {\tt nuflux} by an undocumented normalization factor 5.2 for user-defined fluxes. This difference is explained in the appendix on page~\pageref{app:flux}.

@@

For a built-in neutrino source, one has to specify which
built-in spectrum has to be used, as well as its parameters. The software
will then automatically calculate the neutrino spectrum. Note that in this
case, there is no degree of freedom in the choice of the source units.
The currently available fluxes are described in \Tab~\ref{tab:binfluxes}.
For example, two built-in neutrino factory fluxes are available: $\mu^+$-decay ({\tt @builtin = 1}) and $\mu^-$-decay ({\tt @builtin = 2}). In these cases, the muon energy (enery of the parent particle) has to be specified together with the number of useful decays  per year. Thus, an example to set up a neutrino factory flux is 
\index{aedl}{nuflux@{\tt nuflux}!builtin@{\tt "@builtin}}
\index{aedl}{nuflux@{\tt nuflux}!parent energy@{\tt "@parent\_energy}}
\index{aedl}{nuflux@{\tt nuflux}!stored muons@{\tt "@stored\_muons}}
\index{aedl}{nuflux@{\tt nuflux}!end point@{\tt "@end\_point}}
\index{aedl}{nuflux@{\tt nuflux}!stored ions@{\tt "@stored\_ions}}
\index{aedl}{nuflux@{\tt nuflux}!gamma@{\tt "@gamma}}
\begin{quote}
{\tt nuflux(\#mu\_plus)<\\
\tb  @builtin = 1\\
\tb  @parent\_energy = 50.0\\
\tb  @stored\_muons = 5.33e+20\\
\tb  @time = 8.0\\
>}
\end{quote}
MR: SOME MORE ELABORATION AND AN EXAMPLE FOR THE BETA BEAMS MAY BE APPROPRIATE HERE [WW].
%
\index{aedl}{nuflux@{\tt nuflux}!power@{\tt "@power}}
\index{aedl}{nuflux@{\tt nuflux}!norm@{\tt "@norm}}
\index{aedl}{nuflux@{\tt nuflux}!flux file@{\tt "@flux\_file}}
For a user-defined flux, one has to give it the file name:
\begin{quote}
{\tt nuflux(\#user)<}\\
{\tt \tb @flux\_file = "user\_file\_1.dat"\\
\tb @time = 2.0\\
\tb @power = 4.0\\
\tb @norm = 1e+8}\\
{\tt >}
\end{quote}
In this case, the {\tt @norm} variable is an overall normalization which defines a conversion factor from the fluxes in the file to the units in \GLOBES . In general, there are many ways to give the source power of a 
neutrino source, such as neutrinos per proton on target per area per time frame. Right now, each flux has its own normalization factor, which is
not always straightfoward to calculate. Often, one has to take into account
many things, such as the number of target particles per unit mass. 
In addition, the fluxes will be rescaled by $1/L^2$, which means that the
normalization must contain a factor $L_0^2$. Here $L_0$ is the distance from the source for which the flux is given to the actual neutrino production region. At the end, it is left to the user to ensure that the 
numbers in the flux file give, after the multiplication with {\tt @norm}, 
the proper numbers of produced neutrinos corresponding to the chosen target power {\tt @power}. Usually this adjustment of {\tt @norm} is performed by comparison with known energy spectra for a specific experiment.
For more details on the flux definition, see appendix on page~\pageref{app:flux}.

\index{aedl}{nuflux@{\tt nuflux}!flux file@{\tt "@flux\_file}}\index{norm}{Flux!file}
The software assumes that the given flux file has seven columns and
501 lines with equidistant energies. The format is:
\begin{quotation}
$ E\quad
\Phi_{\nu_e}\quad
\Phi_{\nu_\mu}\quad
\Phi_{\nu_\tau}\quad
\Phi_{\bar\nu_e}\quad
\Phi_{\bar\nu_\mu}\quad
\Phi_{\bar\nu_\tau}$
\end{quotation}
In order to access fluxes at arbitrary energies, linear interpolation 
is used by the software. In general, it is advisable to provide the flux between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}), since these values are used by the software. However, if the energy leaves the range of values given in the file, zero is returned.
The the columns for the fluxes for unused flavours have to be filled all the 
same,\eg\ with zeros.

The flux files accept one-line comments, which start
with {\tt \#} and end with the linefeed character `\verb+\n+', they are
not counted as a line and their content is discarded. These comments
are useful to provide meta information about the fluxes such as units
or the origin of the information. This is also the default method
to point the user to the references one should to cite when using
a particular flux.\index{norm}{Flux!file!comments in}
\index{norm}{Referencing!flux data}

@@

The baseline and matter density profile determine, besides energy and
involved flavors, the neutrino oscillation physics at the experiment
description level. All of the neutrino oscillation parameters are defined at running time.

@@

Furthermore, the matter density profile along the baseline
has to be specified. The simplest matter density profile is a constant matter density profile with the average matter density from the PREM~\cite{Stacey} onion shell model of the earth:\index{norm}{PREM| \see{Matter density}}
\index{norm}{Matter density!of the earth}
\begin{quote}
{\tt \$profiletype=1 }
\end{quote}
%
If your using this option please cite reference~\cite{Stacey}\index{norm}{Referencing!matter profile data}.

@@

Cross sections in \GLOBES\ are given as total cross section divided by energy:
\begin{equation}
\hat\sigma(E)=\sigma(E)/E\,\left[ 10^{-38}\,
\frac{\mathrm{cm}^2}{\mathrm{GeV}} \right]
\end{equation}
The software assumes that the cross section files are text files with 
seven columns and $1001$ lines of the form
\index{norm}{Cross section!file}
\index{aedl}{cross@{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quotation}
$\mathrm{log}_{10} E\quad
\hat\sigma_{\nu_e}\quad
\hat\sigma_{\nu_\mu}\quad
\hat\sigma_{\nu_\tau}\quad
\hat\sigma_{\bar\nu_e}\quad
\hat\sigma_{\bar\nu_\mu}\quad
\hat\sigma_{\bar\nu_\tau}$
\end{quotation}
Here the logarithms of the energy values have to be equidistant. For 
arbitrary energies, linear interpolation is used. If the energy leaves the
range of values given in the file, $0.0$ will be assumed. In general, it is advisable to provide the cross sections in the range between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}).
Unused cross sections have to be filled with zeros, and can not
 be just omitted.

Like the flux files, the cross section files 
accept one-line comments, which start
with {\tt \#} and end with the linefeed character `\verb+\n+', they are
not counted as a line and their content is discarded. This comments
are useful to provide meta information about the cross sections like units
or the origin of the information. This is also the default method
to point the user to the references he/she should to cite when using
a particular cross section.\index{norm}{Cross section!file!comments in}
\index{norm}{Referencing!cross section data}

@@

\index{aedl}{channel@{\tt channel}|(}
Channels\index{norm}{Channel} in \GLOBES\ represent an intermediate level 
between the pure oscillation physics given by the oscillation probability
$P_{\alpha\beta}$, and the total event rates composed of signal and 
background. A channel describes the path from one
initial neutrino flavor in the source to the event rates in the detector for one specific interaction type (IT) and final flavor.  
Therefore, a channel contains the description of the 
initial neutrino flavor, its CP eigenvalue 
(neutrino or antineutrino)\footnote{Currently \GLOBES\ does not support lepton
number violating transitions, \ie\ no transitions from neutrino to 
antineutrino (or vice versa) are considered.}, 
the detected neutrino flavor, the interaction cross sections for the chosen interaction type, and the energy resolution function of the detector.

Before we come to the definition of the channel in AEDL, we introduce the general concept for the calculation of event rates. The first step is to
compute the number of events for each IT in the detector for each 
initial and final neutrino flavor and energy bin. The second step is to include the detector effects coming from the insufficient knowledge in the event reconstruction. These two steps combined lead to the differential event rate spectrum for each initial and final flavor and IT as seen by the detector, which we call the ``channel''. In this section, we focus on the
 first step, \ie , we discuss the definition of the energy resolution function in the next section, since this is a rather comprehensive issue.

The differential event rate for each channel is given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{eqnarray}
\label{eq:master_event}
\frac{dn_{\beta}^{\text{IT}}}{dE'}=&&N\,\int\limits_0^\infty \int\limits_0^\infty dE\,d\hat{E}\quad
\underbrace{\Phi_{\alpha} (E)}_{\mathrm{Production}} \times \nonumber\\
&&\underbrace{\frac{1}{L^2} P_{(\alpha\rightarrow\beta)}(E,L,\rho;\theta_{23},
\theta_{12},\theta_{13},
\Delta m^2_{31},\Delta m^2_{21},\deltacp)}_{\mathrm{Propagation}}
\times \nonumber \\ &&\underbrace{\sigma^{\text{IT}}_f(E)
k_f^{\text{IT}}(E-\hat{E})}_{\mathrm{Interaction}} \times \nonumber \\
&&\underbrace{ T_f(\hat{E}) V_f(\hat{E}-E')}_{\mathrm{Detection}}\,,
\end{eqnarray}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\alpha$ is the initial flavor of the neutrino, 
$\beta$ is the final flavor, $\Phi_{\alpha} (E)$ is the flux of the 
initial flavor at the
source, $L$ is the baseline length, $N$ is a normalization factor, and 
$\rho$ is the matter density. The energies in this formula are given as follows:
\begin{itemize}
\item
 $E$ is the incident neutrino energy, \ie, the actual energy of the 
incoming neutrino (which is not directly accessible to the experiment)
\item
 $\hat{E}$ is the energy of the secondary particle
\item
 $E'$ is the reconstructed neutrino energy, \ie, the measured
neutrino energy as obtained from the experiment
\end{itemize}
The interaction term is composed of 
two factors, which are the total cross section 
$\sigma^{\text{IT}}_\beta(E)$ for the flavor $f$ and
the interaction type IT, and the energy distribution of the 
secondary particle $k_\beta^{\text{IT}}(E-\hat{E})$.
The detector properties are 
modeled by the threshold function $T_\beta(\hat{E})$, coming from the the 
limited resolution or the cuts in the analysis, and the energy resolution 
function $V_\beta(\hat{E}-E')$ of the secondary particle.

Since it is a lot of effort to solve this double integral numerically,
we split up the two integrations. First, we evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$. We define:
\begin{eqnarray}
\label{eq:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is also often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix\index{norm}{Energy!resolution}\index{norm}{Smear matrix}. The function $\epsilon_\beta^{\text{IT}}(E')$ will later be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. The detailed definition and initialization of the energy resolution function is described in \Sec~\ref{sec:energy}.

@@

A simple channel definition in \GLOBES\ consists of the flux,
the CP-sign of the initial state, the initial flavor, the final flavor,
the cross sections, and the energy resolution function. In order to refer to
flux, cross sections, and energy resolution functions, they have to be 
defined before with their {\tt \#name} in the respective environments. 
Thus, a simple definition of a channel\index{aedl}{channel@{\tt channel}} is
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: muon: muon: \#cross: \#energy\\
>}
\end{quote}
%
\index{norm}{Oscillation!switching off}
It is also possible to define a channel as no-oscillation by using the
prefix {\tt NOSC\_}\index{aedl}{channel@{\tt channel}!{\tt NOSC\_}} 
in either the initial  flavour or 
the final flavour, like this
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: NOSC\_muon: muon: \#cross: \#energy\\
>}
\end{quote}
%
In this case all diagonal probabilities $P_{\alpha\alpha}$ are unity, and all off-diagonal probablities  $P_{\alpha\beta}$ are zero. This is,
for instance, useful for
neutral current events, since these do not depend on any oscillation 
parameters\footnote{At least in the absence of sterile neutrinos}. The channels
marked as {\tt NOSC\_} are already computed by \GLB{SetRates}
and do not have to be recomputed in the subsequent fit (which calls the undocumented function \GLB{SetNewRates}).
Therefore this feature can be used to speed up the rate 
computation considerably, especially in cases where a large set
of channels exist which are only used for the computation of backgrounds.
Usually it is an excellent approximation to treat backgrounds as if they
were not affected by oscillations\footnote{In the case that the backgrounds have
a sizeable dependence on the oscillation parameters, they carry information
on the oscillation parameters, and therefore behave more like a signal.}.

@@

In this section, we first discuss the principles of the energy smearing, where it is assumed that the reader is familiar with \Sec~\ref{sec:channel}. Then we introduce an automatic energy smearing algorithm, which is fairly simple to understand and applicable to most beam-based experiments. 
In most cases, the reader may want to continue to the next section after reading these two subsections. In the third subsection, we describe a more elaborate (and slower) smearing algorithm, which can be used together with rather fast neutrino oscillations compared to the bin size, such as for
(solar) reactor experiments to avoid aliasing effects. Eventually, we show how one can use a manual smearing matrix instead of using one of the implemented algorithms.

@@

The energy resolution function $R^c(E,E')$ has been already introduced in 
\Sec~\ref{sec:channel}, where a definition
has been given in \eq~(\ref{eq:e_res}). Instead of using
 \eq~(\ref{eq:e_res}) directly, we apply a slightly different
definition of the post-smearing efficiencies $\epsilon(E')$. 
In general, $\epsilon(E')$ has to be
determined by means of a Monte Carlo simulation of the experiment. 
This usually involves a binning of the simulated events in the 
reconstructed energy $E'$. Therefore, one can simplify \eq~(\ref{eq:events_bin}) by
\begin{equation}
\label{eq:post_smearing}
\int_{E_i-\Delta E_i/2}^{E_i+\Delta E_i/2} dE' \, R^c(E,E') \, \epsilon^c(E')
 \simeq \hat\epsilon_i^c \cdot \int_{E_i-\Delta E_i/2}^{E_i+\Delta E_i/2} dE' 
\quad R^c(E,E')\,.
\end{equation}
Here the  $\hat\epsilon_i^c$ are the 
binned ``post-smearing'' efficiencies, which will be set within the corresponding {\tt channel} environment (see below).
From \eq~(\ref{eq:events_bin}) it is obvious that the integration with respect to the reconstructed energy $E'$ can be
performed independently of the oscillation parameters. We define
the ``bin kernel'' $K_i^c$ for the $i$th bin as
\begin{equation}
\label{eq:kernel}
K_i^c(E) \equiv \int_{E_i-\Delta E_i/2}^{E_i+\Delta E_i/2} dE' 
\quad R^c(E,E')\,.
\end{equation}
With this definition, \eq~(\ref{eq:events_bin}) can be re-written as
\begin{equation}
\label{eq:simple_int}
n_i^c=N/L^2 \,
\hat\epsilon_i^c \, \int\limits_0^\infty dE\quad  \underbrace{\Phi^c(E)\,
P^c(E)\,
\sigma^c(E) \, K_i^c(E)\,}_{f(E)}. 
\end{equation}

@@

In some cases, one may use the output of a detector Monte Carlo simulation
directly. Then one can use ''manual'' energy smearing instead  of the
automatic energy smearing algorithms.

>>>>

In this chapter, we give a detailed description of the \AEDL\ features. We also show the underlying mathematical concepts where applicable. We do not exactly follow the separation of source, oscillation, and detection properties, since most issues more or less involve the detection. Instead,
we illustrate many of the features of the \GLOBES\ simulation successively
in the logical order of their definition, and demonstrate how they translate into \AEDL .

@@

As we have disussed before, \GLOBES\ can only deal with point sources. Thus,  it is not possible to study effects of the finite size of the neutrino production region, such as in the sun or in reactor experiments with many
neutrino sources (\eg, KamLAND). Therefore, a neutrino source in \GLOBES\ can, in general, be characterized by the flux spectrum for each neutrino flavor, the CP sign (neutrinos or antineutrinos), and the total luminosity
of the source.

@@

\index{aedl}{nuflux@{\tt nuflux}!time@{\tt "@time}}
There are two principal ways to initialize a neutrino flux: 
Either one can use 
a built-in source, or one can provide a file. In both cases,
a flux is defined by the environment {\tt nuflux}, such as
\index{aedl}{nuflux@{\tt nuflux}}
\begin{quote}
  {\tt nuflux(\#name)<\\
\tb $\ldots$\\
\tb @time = 8.0 \\
>}
\end{quote}
with a running time of $8$ years. Note that the running time is used within
the {\tt nuflux} environment. This feature can be used to load the neutrino
and antineutrino fluxes in an accelerator experiment separately, in order
to combine them with different running times in the respective operation modes.
The name of the flux {\tt \#name} will later be referred to in the channel definitions. Note that \GLOBES\ versions older than 3.0 use the still supported {\tt flux} environment, which is different from {\tt nuflux} by an undocumented normalization factor 5.2 for user-defined fluxes. This difference is explained in the appendix on page~\pageref{app:flux}.

@@

For a built-in neutrino source, one has to specify which
built-in spectrum should be used, as well as its parameters. The software
will then automatically calculate the neutrino spectrum. Note that in this
case, there is no degree of freedom in the choice of the source units.
The currently available fluxes are described in \Tab~\ref{tab:binfluxes}.
For example, two built-in neutrino factory fluxes are available: $\mu^+$-decay ({\tt @builtin = 1}) and $\mu^-$-decay ({\tt @builtin = 2}). In these cases, the muon energy (enery of the parent particle) has to be specified together with the number of useful decays  per year. Thus, an example to set up a neutrino factory flux is 
\index{aedl}{nuflux@{\tt nuflux}!builtin@{\tt "@builtin}}
\index{aedl}{nuflux@{\tt nuflux}!parent energy@{\tt "@parent\_energy}}
\index{aedl}{nuflux@{\tt nuflux}!stored muons@{\tt "@stored\_muons}}
\index{aedl}{nuflux@{\tt nuflux}!end point@{\tt "@end\_point}}
\index{aedl}{nuflux@{\tt nuflux}!stored ions@{\tt "@stored\_ions}}
\index{aedl}{nuflux@{\tt nuflux}!gamma@{\tt "@gamma}}
\begin{quote}
{\tt nuflux(\#mu\_plus)<\\
\tb  @builtin = 1\\
\tb  @parent\_energy = 50.0\\
\tb  @stored\_muons = 5.33e+20\\
\tb  @time = 8.0\\
>}
\end{quote}
MR: SOME MORE ELABORATION AND AN EXAMPLE FOR THE BETA BEAMS MAY BE APPROPRIATE HERE [WW].
%
\index{aedl}{nuflux@{\tt nuflux}!power@{\tt "@power}}
\index{aedl}{nuflux@{\tt nuflux}!norm@{\tt "@norm}}
\index{aedl}{nuflux@{\tt nuflux}!flux file@{\tt "@flux\_file}}
For a user-defined flux, one has to specify the file name:
\begin{quote}
{\tt nuflux(\#user)<}\\
{\tt \tb @flux\_file = "user\_file\_1.dat"\\
\tb @time = 2.0\\
\tb @power = 4.0\\
\tb @norm = 1e+8}\\
{\tt >}
\end{quote}
In this case, the {\tt @norm} variable is an overall normalization which defines a conversion factor from the fluxes in the file to the units in \GLOBES . In general, there are many ways to give the source power of a 
neutrino source, such as neutrinos per proton on target per area per time frame. Right now, each flux has its own normalization factor, which is
not always straightfoward to calculate. Often, one has to take into account
many things, such as the number of target particles per unit mass. 
In addition, the fluxes will be rescaled by $1/L^2$, which means that the
normalization must contain a factor $L_0^2$. Here $L_0$ is the distance from the source for which the flux is given to the actual neutrino production region. At the end, it is left to the user to ensure that the 
numbers in the flux file give, after the multiplication with {\tt @norm}, 
the proper numbers of produced neutrinos corresponding to the chosen target power {\tt @power}. Usually this adjustment of {\tt @norm} is performed by comparison with known energy spectra for a specific experiment.
For more details on the flux definition, see page~\pageref{app:flux} in the appendix.

\index{aedl}{nuflux@{\tt nuflux}!flux file@{\tt "@flux\_file}}\index{norm}{Flux!file}
The software assumes that the given flux file has seven columns and
501 lines with equidistant energies. The format is:
\begin{quotation}
$ E\quad
\Phi_{\nu_e}\quad
\Phi_{\nu_\mu}\quad
\Phi_{\nu_\tau}\quad
\Phi_{\bar\nu_e}\quad
\Phi_{\bar\nu_\mu}\quad
\Phi_{\bar\nu_\tau}$
\end{quotation}
In order to access fluxes at arbitrary energies, linear interpolation 
is used by \GLOBES. In general, it is advisable to provide the flux
between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}),
since this is the energy range considered in the simulation. However,
if part of this interval is omitted in the flux file, zero will be
used there. If some neutrino flavours are not used in the simulation,
the corresponding columns in the flux file have to be filled nevertheless,
\eg\ with zeros.

The flux files accept one-line comments, which start
with {\tt \#} and end with the linefeed character `\verb+\n+', they are
not counted as a line and their content is discarded. These comments
are useful to provide meta information about the fluxes such as units
or the origin of the information. This is also the default method
to point the user to the references that should be cited when using
a particular flux file.\index{norm}{Flux!file!comments in}
\index{norm}{Referencing!flux data}

@@

Besides the energy and the involved flavors, the neutrino oscillation
physics depends on the baseline and the matter density profile.
All neutrino oscillation parameters are defined at running time.

@@

Furthermore, the matter density profile along the baseline
has to be specified. The simplest profile is a constant matter density equal to the average matter density from the PREM~\cite{Stacey} onion shell model of the earth:\index{norm}{PREM| \see{Matter density}}
\index{norm}{Matter density!of the earth}
\begin{quote}
{\tt \$profiletype=1 }
\end{quote}
%
If you are using this option please cite reference~\cite{Stacey}\index{norm}{Referencing!matter profile data}.

@@

Cross sections in \GLOBES\ are given as total cross section divided by energy:
\begin{equation}
\hat\sigma(E)=\sigma(E)/E\,\left[ 10^{-38}\,
\frac{\mathrm{cm}^2}{\mathrm{GeV}} \right]
\end{equation}
The software assumes that the cross section files are text files with 
seven columns and $1001$ lines of the form
\index{norm}{Cross section!file}
\index{aedl}{cross@{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quotation}
$\mathrm{log}_{10} E\quad
\hat\sigma_{\nu_e}\quad
\hat\sigma_{\nu_\mu}\quad
\hat\sigma_{\nu_\tau}\quad
\hat\sigma_{\bar\nu_e}\quad
\hat\sigma_{\bar\nu_\mu}\quad
\hat\sigma_{\bar\nu_\tau}$
\end{quotation}
Here the logarithms of the energy values have to be equidistant. For 
arbitrary energies, linear interpolation is used. If the energy leaves the
range of values given in the file, $0.0$ will be assumed. In general, it is advisable to provide the cross sections in the range between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}).
Cross sections for unused neutrino flavours have to be filled with zeros, and
can not just be omitted.

Like the flux files, the cross section files 
accept one-line comments, which start
with {\tt \#} and end with the linefeed character `\verb+\n+'; they are
not counted as a line and their content is discarded. These comments
are useful to provide meta information about the cross sections like units
or the origin of the information. This is also the default method
to point the user to the references he/she should cite when using
a particular cross section file.\index{norm}{Cross section!file!comments in}
\index{norm}{Referencing!cross section data}

@@

\index{aedl}{channel@{\tt channel}|(}
Channels\index{norm}{Channel} in \GLOBES\ represent an intermediate level 
between the pure oscillation physics given by the oscillation probability
$P_{\alpha\beta}$, and the final event rates composed of signal and 
background. A channel describes the path from one
initial neutrino flavor in the source to the event rates in the detector for one specific interaction type (IT) and final flavor.  
Therefore, a channel contains the description of the 
initial neutrino flavor, its CP eigenvalue 
(neutrino or antineutrino)\footnote{Currently, \GLOBES\ does not support lepton
number violating transitions, \ie\ no transitions from neutrino to 
antineutrino (or vice versa) are considered.}, 
the detected neutrino flavor, the interaction cross sections for the chosen interaction type, and the energy resolution function of the detector.

Before we come to the definition of channels in AEDL, we introduce the general concept for the calculation of event rates. The first step is to
compute the number of events for each IT in the detector for each 
initial and final neutrino flavor and energy bin. The second step is to include the detector effects coming from the insufficient knowledge in the event reconstruction.
The combination of these two steps leads to the differential event rate spectrum for each initial and final flavor and IT as seen by the detector, which we call the ``channel''. In this section, we focus on the
 first step, \ie , we discuss the definition of the energy resolution function in the next section, since this is a rather comprehensive issue.

The differential event rate for each channel is given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{eqnarray}
\label{eq:master_event}
\frac{dn_{\beta}^{\text{IT}}}{dE'}=&&N\,\int\limits_0^\infty \int\limits_0^\infty dE\,d\hat{E}\quad
\underbrace{\Phi_{\alpha} (E)}_{\mathrm{Production}} \times \nonumber\\
&&\underbrace{\frac{1}{L^2} P_{(\alpha\rightarrow\beta)}(E,L,\rho;\theta_{12},
\theta_{13},\theta_{23},
\Delta m^2_{31},\Delta m^2_{21},\deltacp)}_{\mathrm{Propagation}}
\times \nonumber \\ &&\underbrace{\sigma^{\text{IT}}_f(E)
k_f^{\text{IT}}(E-\hat{E})}_{\mathrm{Interaction}} \times \nonumber \\
&&\underbrace{ T_f(\hat{E}) V_f(\hat{E}-E')}_{\mathrm{Detection}}\,,
\end{eqnarray}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\alpha$ is the initial flavor of the neutrino, 
$\beta$ is the final flavor, $\Phi_{\alpha} (E)$ is the flux of the 
initial flavor at the
source, $L$ is the baseline length, $N$ is a normalization factor, and 
$\rho$ is the matter density. The energies in this formula are given as follows:
\begin{itemize}
\item
 $E$ is the incident neutrino energy, \ie, the actual energy of the 
incoming neutrino (which is not directly accessible to the experiment)
\item
 $\hat{E}$ is the energy of the secondary particle
\item
 $E'$ is the reconstructed neutrino energy, \ie, the measured
neutrino energy as obtained from the experiment
\end{itemize}
The interaction term is composed of 
two factors, which are the total cross section 
$\sigma^{\text{IT}}_\beta(E)$ for the flavor $f$ and
the interaction type IT, and the energy distribution of the 
secondary particle $k_\beta^{\text{IT}}(E-\hat{E})$.
The detector properties are 
modeled by the threshold function $T_\beta(\hat{E})$, coming from the the 
limited resolution or the cuts in the analysis, and the energy resolution 
function $V_\beta(\hat{E}-E')$ of the secondary particle.

Since it is computationally very expensive to solve this double integral
numerically, we split up the two integrations. The integral over $\hat{E}$
depends only on the terms containing $\hat{E}$, i.e., on $k_\beta^{\text{IT}}(E-\hat{E})$,
$ T_\beta(\hat{E})$, and $V_\beta(\hat{E}-E')$. These terms do not depend
on the oscillation parameters, so they will not vary during the fit,
and the $\hat{E}$ integral can be pre-computed in the initialization phase.
We define:
\begin{eqnarray}
\label{eq:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix\index{norm}{Energy!resolution}\index{norm}{Smear matrix}. The function $\epsilon_\beta^{\text{IT}}(E')$ will later be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. The detailed definition and initialization of the energy resolution function is described in \Sec~\ref{sec:energy}.

@@

A simple channel definition in \GLOBES\ consists of the flux,
the CP-sign of the initial state, the initial flavor, the final flavor,
the cross sections, and the energy resolution function. In order to refer to
the fluxes, cross sections, and energy resolution functions, they have to be 
defined first with their {\tt \#name} in the respective environments. 
A simple definition of a channel\index{aedl}{channel@{\tt channel}} is
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: muon: muon: \#cross: \#energy\\
>}
\end{quote}
%
\index{norm}{Oscillation!switching off}
It is also possible to define a channel as no-oscillation by using the
prefix {\tt NOSC\_}\index{aedl}{channel@{\tt channel}!{\tt NOSC\_}} 
in either the initial  flavour or 
the final flavour, like this
\begin{quote}
{\tt channel(\#channel\_1)<\\
\tb @channel = \#flux : $+$: NOSC\_muon: muon: \#cross: \#energy\\
>}
\end{quote}
%
In this case all diagonal probabilities $P_{\alpha\alpha}$ are unity, and all off-diagonal probablities  $P_{\alpha\beta}$ are zero. This is,
for instance, useful for
neutral current events, since these do not depend on any oscillation 
parameters\footnote{At least in the absence of sterile neutrinos}. The channels
marked as {\tt NOSC\_} are already computed by \GLB{SetRates}
and do not have to be recomputed in the subsequent fit (which calls the undocumented function \GLB{SetNewRates}).
Therefore this feature can be used to speed up the rate 
computation considerably, especially in cases where a large set
of channels exist which are only used for the computation of backgrounds.
Usually, it is an excellent approximation to treat backgrounds as if they
were not affected by oscillations\footnote{In the case that the backgrounds have
a sizeable dependence on the oscillation parameters, they carry information
on the oscillation parameters, and therefore behave more like a signal.}.

@@

In this section, we first discuss the principles of the energy smearing, where it is assumed that the reader is familiar with \Sec~\ref{sec:channel}. Then we introduce an automatic energy smearing algorithm, which is fairly simple to understand and applicable to most beam-based experiments. 
In most cases, the reader may want to proceed to the next section after reading these two subsections. In the third subsection, we describe a more elaborate (and slower) smearing algorithm, which is useful to avoid aliasing effects if the neutrino oscillations are rather fast compared to the bin size, as is the case for
(solar) reactor experiments. Eventually, we show how one can use a manual smearing matrix instead of using one of the implemented algorithms.

@@

The energy resolution function $R^c(E,E')$ and the post-smearing
efficiencies $\epsilon(E')$ have already been introduced in 
\Sec~\ref{sec:channel}, where a definition
has been given in \eq~(\ref{eq:e_res}). Instead of using
 \eq~(\ref{eq:e_res}) directly, we apply a slightly different
definition of the post-smearing efficiencies $\epsilon(E')$. 
In general, they have to be
determined by means of a Monte Carlo simulation of the experiment. 
This usually involves a binning of the simulated events in the 
reconstructed energy $E'$. Therefore, one can simplify \eq~(\ref{eq:events_bin}) by
\begin{equation}
\label{eq:post_smearing}
\int_{E_i-\Delta E_i/2}^{E_i+\Delta E_i/2} dE' \, R^c(E,E') \, \epsilon^c(E')
 \simeq \hat\epsilon_i^c \cdot \int_{E_i-\Delta E_i/2}^{E_i+\Delta E_i/2} dE' 
\quad R^c(E,E')\,.
\end{equation}
Here the  $\hat\epsilon_i^c$ are the 
binned ``post-smearing'' efficiencies, which will be set within the corresponding {\tt channel} environment (see below).
From \eq~(\ref{eq:events_bin}) it is obvious that the integration with respect to the reconstructed energy $E'$ can be
performed independently of the oscillation parameters. We define
the ``bin kernel'' $K_i^c$ for the $i$th bin as
\begin{equation}
\label{eq:kernel}
K_i^c(E) \equiv \int_{E_i-\Delta E_i/2}^{E_i+\Delta E_i/2} dE' 
\quad R^c(E,E')\,.
\end{equation}
With this definition, \eq~(\ref{eq:events_bin}) can be re-written as
\begin{equation}
\label{eq:simple_int}
n_i^c=N/L^2 \,
\hat\epsilon_i^c \, \int\limits_0^\infty dE\quad  \underbrace{\Phi^c(E)\,
P^c(E)\,
\sigma^c(E) \, K_i^c(E)\,}_{f(E)}. 
\end{equation}

@@

In some cases, one may want to use the output of a detector Monte Carlo simulation
directly. Then one can use ''manual'' energy smearing instead  of the
automatic energy smearing algorithms.
